---
author: "Panagiotis Anastasakis"
output: pdf_document
header-includes:
  - |
    ```{=latex}
    \usepackage{fvextra}
    \usepackage{amsmath}
    \usepackage{algpseudocode}
    \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
      showspaces = false,
      showtabs = false,
      breaklines,
      commandchars=\\\{\}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(28)
```


## Motivating example for GUIDE


```{r}

library(ggplot2)
library(pheatmap)
library(grid)

```



```{r}

n = 3000
p = 200
k = 20

noise.sd.1 = 0.3
noise.sd.2 = 0.5

block.list = block_sim(n, p, k, lower = -2, upper = 2, dependence.prob = 0.1,
                       noise.sd.1 = noise.sd.1, noise.sd.2 = noise.sd.2)

W = block.list$W

W.xl = block.list$W.xl
W.lt = block.list$W.lt

weights.true = rbind(W.xl, W.lt)



guide.list = get_guide(W, K = k, ica_runs = 1, verbose=F) ## no need for ICASSO in simple optimization settings, so set ica_runs=1
W.xl.guide = guide.list$W.xl
W.lt.guide = guide.list$W.lt
weights.guide = rbind(W.xl.guide, W.lt.guide)

weights.guide.matched = get_matched_weights(weights.true = weights.true,
                                            weights.estimated = weights.guide,
                                            n.variants = n)

W.xl.guide.matched = weights.guide.matched$weights.matched.xl
W.lt.guide.matched = weights.guide.matched$weights.matched.lt


degas.list = get_tsvd(scale(W), K = k)
W.xl.degas = degas.list$U
W.lt.degas = degas.list$V
weights.degas = rbind(W.xl.degas, W.lt.degas)

weights.degas.matched = get_matched_weights(weights.true = weights.true,
                                            weights.estimated = weights.degas,
                                            n.variants = n)

W.xl.degas.matched = weights.degas.matched$weights.matched.xl
W.lt.degas.matched = weights.degas.matched$weights.matched.lt


## GUIDE

## save at 12.5, 9
plot_simulation(W.xl, W.xl.guide.matched, xlab = "GUIDE Variant Weights", ylab = "Simulated Variant Weights", main = "GUIDE & Simulated Variant Weights")
plot_simulation(W.lt, W.lt.guide.matched, xlab = "GUIDE Trait Weights", ylab = "Simulated Trait Weights", main = "GUIDE & Simulated Trait Weights")


#### TSVD

## save at 12.5, 9
plot_simulation(W.xl, W.xl.degas.matched, xlab = "DeGAs Variant Weights", ylab = "Simulated Variant Weights", main = "DeGAs & Simulated Variant Weights")
plot_simulation(W.lt, W.lt.degas.matched, xlab = "DeGAs Trait Weights", ylab = "Simulated Trait Weights", main = "DeGAs & Simulated Trait Weights")


```


### Showcasing how GUIDE's approach for estimating a number of components can become incosistent with small genetic architectures


```{r}

set.seed(22)


n.1 = 200
p.1 = 30

n.2 = 1000
p.2 = 100

n.blocks.1 = 5
n.blocks.2 = 10

K.1 = 10
K.2 = 30

nlat.output.1 = c()
nlat.output.2 = c()

n.reps = 100

for (ii in 1:n.reps) {
  
  block.list = block_sim(n.1, p.1, n.blocks.1, lower = -2, upper = 2, dependence.prob = 0,
                         noise.sd.1 = 0.1, noise.sd.2 = 0.5)

  W = block.list$W
  
  K.est.1 = get_nlatents(W, starting.K = K.1, validation.reps = 30, alg.typ = "deflation")$K
  
  nlat.output.1 = c(nlat.output.1, K.est.1)
  
  print(paste("Simulation", ii, "completed."))
}

ggplot(data.frame(nlat.output.1), aes(x = nlat.output.1)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue") +
  labs(title = "GUIDE Component Estimation, Structure 1",
       x = "Estimated Components",
       y = "Count" ) + 
  theme_minimal() +
  scale_x_continuous(breaks = 5:10) + 
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 18),  
        axis.text.y = element_text(size = 18),
        plot.margin = margin(20, 10, 20, 15))


for (ii in 1:n.reps) {
  
  block.list = block_sim(n.2, p.2, n.blocks.2, lower = -2, upper = 2, dependence.prob = 0,
                         noise.sd.1 = 0.1, noise.sd.2 = 0.5)

  W = block.list$W
  
  K.est.2 = get_nlatents(W, starting.K = K.2, validation.reps = 30, alg.typ = "deflation")$K
  
  nlat.output.2 = c(nlat.output.2, K.est.2)
  
  print(paste("Simulation", ii, "completed."))
}


ggplot(data.frame(nlat.output.2), aes(x = nlat.output.2)) +
  geom_histogram(binwidth = 0.5, fill = "lightblue") +
  labs(title = "GUIDE Component Estimation, Structure 2",
       x = "Estimated Components",
       y = "Count" ) + 
  theme_minimal() +
  scale_x_continuous(breaks = 10:11) +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 18),  
        axis.text.y = element_text(size = 18),
        plot.margin = margin(20, 10, 20, 15))




```



### GUIDE is mostly correct with the larger genetic architecture, but sometimes is wrong despite the clear present of 10 non-random components


```{r}


block.list = block_sim(n.2, p.2, n.blocks.2, lower = -2, upper = 2, dependence.prob = 0,
                         noise.sd.1 = 0.1, noise.sd.2 = 0.5)

W = block.list$W


d = get_tsvd(scale(W), K = p.2)$d

ggplot(data.frame(index = 1:length(d), d = d), aes(x = index, y = d)) +
  geom_point(color = "black", size = 2) +       
  labs(title = "Scree plot for SVD on Structure 2",
       x = "Component Number", 
       y = "Singular Value") +
  theme_minimal() +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 16),  
        axis.text.y = element_text(size = 16),
        plot.margin = margin(20, 10, 20, 15))


```


### GUIDE's estimation of the number of components also breaks down with increasing levels of component dependence


```{r}

n = 1000
p = 100
n.blocks = 10

K = 30

n.reps = 100

dependence_probs = c(0.05, 0.1, 0.15, 0.2)


nlat.outputs = matrix(0, nrow = length(dependence_probs), ncol = n.reps) ## each row is one dependence probability

for (jj in 1:n.reps) {
  
  for (ii in 1:length(dependence_probs)){
    
    block.list = block_sim(n, p, n.blocks, lower = -2, upper = 2,
                           dependence.prob = dependence_probs[ii],
                           noise.sd.1 = 0.1, noise.sd.2 = 0.5)

    W = block.list$W
    
    nlat.output = get_nlatents(W, starting.K = K, validation.reps = 30, alg.typ = "deflation")$K
    
    nlat.outputs[ii, jj] = nlat.output
  }
  
  print(paste("Simulation", jj, "completed."))
}



ggplot(data.frame(nlat.outputs[1,]), aes(x = nlat.outputs[1,])) +
  geom_histogram(binwidth = 0.5, fill = "thistle") +
  labs(title = expression(paste("GUIDE Component Estimation, ", italic(P)[dep], " = 0.05")),
       x = "Estimated Components",
       y = "Count" ) +
  scale_x_continuous(breaks = 9:11) +
  theme_minimal() +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 16),  
        axis.text.y = element_text(size = 16),
        plot.margin = margin(20, 10, 20, 15))


ggplot(data.frame(nlat.outputs[2,]), aes(x = nlat.outputs[2,])) +
  geom_histogram(binwidth = 0.5, fill = "thistle") +
  labs(title = expression(paste("GUIDE Component Estimation, ", italic(P)[dep], " = 0.1")),
       x = "Estimated Components",
       y = "Count" ) +
  theme_minimal() +
  scale_x_continuous(breaks = 8:10) +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 16),  
        axis.text.y = element_text(size = 16),
        plot.margin = margin(20, 10, 20, 15))

ggplot(data.frame(nlat.outputs[3,]), aes(x = nlat.outputs[3,])) +
  geom_histogram(binwidth = 0.5, fill = "thistle") +
  labs(title = expression(paste("GUIDE Component Estimation, ", italic(P)[dep], " = 0.15")),
       x = "Estimated Components",
       y = "Count" ) +
  theme_minimal() +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 16),  
        axis.text.y = element_text(size = 16),
        plot.margin = margin(20, 10, 20, 15))


ggplot(data.frame(nlat.outputs[4,]), aes(x = nlat.outputs[4,])) +
  geom_histogram(binwidth = 0.5, fill = "thistle") +
  labs(title = expression(paste("GUIDE Component Estimation, ", italic(P)[dep], " = 0.2")),
       x = "Estimated Components",
       y = "Count" ) +
  theme_minimal() +
  scale_x_continuous(breaks = 2:7) +
  theme(plot.title = element_text(size = 28, hjust = 0.5, vjust = 3),
        axis.title.x = element_text(size = 23, vjust = -1),              
        axis.title.y = element_text(size = 23, vjust = 3),
        axis.text.x = element_text(size = 16),  
        axis.text.y = element_text(size = 16),
        plot.margin = margin(20, 10, 20, 15))


```

